Linear response theory and optimal control for a molecular system under nonequilibrium conditions

### 2013-08-27 writedown/linear.response/paper.tex
remove all red things

\paragraph*{Remark.}
\recheck{Ciccotti and Jacucci have pointed out that
  the equilibrium linear response
can also be calculated by a direct nonequilibrium simulation of the
perturbed process, and  the nonequilibrium
approach has a smaller statistical uncertainty in comparison with the
standard Green-Kubo fomula~\cite{ciccotti1975direct}.
}

@@

In principle, optimal control problems such as (\ref{ocp1})--(\ref{ocp2}) can be solved by dynamic programming, i.e.~by solving the corresponding Hamilton-Jacobi-Bellman PDE \cite{fleming2006}. Except for very simple, essentially one-dimensional systems, solving Hamilton-Jacobi-Bellman equations is not an easy task, so we pursue a different strategy here. The idea is to use that we have restricted our space of admissible controls to functions of the form (\ref{admissible}) with given basis vector fields and that we can do a gradient search in the unknown coefficients $a_{k}$, using the iteration
\[
u^{(n+1)} = u^{(n)} + \tau_{n}\nabla I(u^{(n)})\,,
\]
with $\tau_{n}>0$ being an adjustable parameter that determines the length of each gradient step. The gradient of $I$ can be easily evaluated using the linear response formula. To see this recall the notion of functional (G\^ateaux) derivatives as directional derivatives along a function $v$ (from a suitable function space): 
\begin{equation*}
\frac{\delta I}{\delta u} = \left.\frac{d}{d\eps}\right|_{\eps=0} I(u+\eps v) = \langle\nabla I(u),v\rangle\,.
\end{equation*} 
For existence of the functional derivative \wrt general It\^o stochastic processes, we refer to \cite{bell2006}.  
Now the idea is to use that we have restricted our space of admissible controls to functions of the form (\ref{admissible}) with given basis vector fields and do a gradient search in the unknown coefficients $a_{k}$. This requires to compute the gradient \wrt the coefficients. Let the vector  
\[
\delta u = \sum_{k=1}^K \delta a_k \Phi_k(t)
\]   
denote the direction along which we want to differentiate where $\delta a_{1},\ldots,\delta a_{K}$ are the coefficients of the vector $\delta u$ in the basis of the $\Phi_{k}$, and note that 
\begin{equation*}
\begin{aligned}
\frac{\delta I}{\delta u} & = \left.\frac{d}{d\eps}\right|_{\eps=0} I(u+\eps \delta u)\\
& = \left(\lim_{\eps\to 0}\frac{I(u+\eps\delta u) - I(u)}{\eps}\right)\cdot \delta u\\
& = \left(\lim_{\eps\to 0}\frac{\bE_{P^{\eps}}[\psi(u+\eps \delta u)] - \bE_{P^{0}}[\psi(u)]}{\eps}\right)\cdot\delta u\\
\end{aligned}
\end{equation*} 
provided that the limit exists. The above iteration therefore is equivalent to
\begin{equation}\label{gd}
a_{k}^{(n+1)} = a_{k}^{(n)} + \tau_{n}\left.\frac{\partial I}{\partial a_{k}}\right|_{a_{k}=a_{k}^{(n)}}\,,
\end{equation}
with
\begin{equation}\label{dida}
\frac{\partial I}{\partial a_{k}} = \bE_{P^0}\left[\psi\int_{0}^{T}(\sigma^{-1}D(q_{s})\Phi_{k}(s))\cdot dB_{s} + \int_{0}^{T}u(s)\cdot\Phi_{k}(s)\,ds \right] \,,
\end{equation}
and $\bE_{P^0}[\cdot]$ being the expectation over all realizations of (\ref{eqn:langevin-1d}) with $\eps=0$ and $u=u^{(n)}$ being the current iterate and arbitrary initial conditions $x=(q,p)$, e.g., 
\[
\bE_{P^{0}}[\psi] = \int_{\R^{2n}}\bE_{P^{0}_{x}}[\psi]\,d\mu(x) \,,
\]
with $P_{x}^{0}$ denoting the corresponding path measure with deterministic initial condition, and $\mu$ is the prescribed initial distribution.
\recheck{A related result on computing the parameter sensitivity can be found
  in Ref.~\cite{warren2012malliavin}.}
Note that we can compute all partial derivatives $\partial I/\partial a_k$, $k=1,\ldots,K$ from just one ensemble of nonequilibrium paths of $(q_{t},p_{t})$. Further note that, in general, the updated coefficient $a^{(n+1)}$ and hence the updated control $u^{(n+1)}$ will depend on the distribution of the initial conditions at the $n$-th iteration stage; in particular, if the controls are computed on-the-fly, the controls will actually be feedback controls depending on the current state and the time elapsed since $t=0$.

@@

\recheck{
  We derived the first and second order
  response formulas for nonequilibrium molecular
dynamics (driven Langevin dynamics), which is a generalization of the
standard finite-time equilibrium linear response theory.
The novel nonequilibrium response theory
does not rely on any reversibility or stationarity assumption on the dynamics.
We validated the formula in numerical
experiments in comparison to brute-force nonequilibrium
simulations. There, we demonstrated that the nonequilibrium linear
response formula allows to extend the algorithmic use of linear
response theory to significantly stronger perturbations of the system
since it permits intermediate drivings, with which the effective perturbations
still lie in the linear range.
}

>>>>

\paragraph*{Remark.}
{Ciccotti and Jacucci have pointed out that
  the equilibrium linear response
can also be calculated by a direct nonequilibrium simulation of the
perturbed process, and  this nonequilibrium
approach has a smaller statistical uncertainty in comparison with the
standard Green-Kubo fomula~\cite{ciccotti1975direct}.
}

@@

In principle, optimal control problems such as (\ref{ocp1})--(\ref{ocp2}) can be solved by dynamic programming, i.e.~by solving the corresponding Hamilton-Jacobi-Bellman PDE \cite{fleming2006}. Except for very simple, essentially one-dimensional systems, solving Hamilton-Jacobi-Bellman equations is not an easy task, so we pursue a different strategy here. The idea is to use that we have restricted our space of admissible controls to functions of the form (\ref{admissible}) with given basis vector fields and that we can do a gradient search in the unknown coefficients $a_{k}$, using the iteration
\[
u^{(n+1)} = u^{(n)} + \tau_{n}\nabla I(u^{(n)})\,,
\]
with $\tau_{n}>0$ being an adjustable parameter that determines the length of each gradient step. The gradient of $I$ can be easily evaluated using the linear response formula. To see this recall the notion of functional (G\^ateaux) derivatives as directional derivatives along a function $v$ (from a suitable function space): 
\begin{equation*}
\frac{\delta I}{\delta u} = \left.\frac{d}{d\eps}\right|_{\eps=0} I(u+\eps v) = \langle\nabla I(u),v\rangle\,.
\end{equation*} 
For existence of the functional derivative \wrt general It\^o stochastic processes, we refer to \cite{bell2006}.  
Now the idea is to use that we have restricted our space of admissible controls to functions of the form (\ref{admissible}) with given basis vector fields and do a gradient search in the unknown coefficients $a_{k}$. This requires to compute the gradient \wrt the coefficients. Let the vector  
\[
\delta u = \sum_{k=1}^K \delta a_k \Phi_k(t)
\]   
denote the direction along which we want to differentiate where $\delta a_{1},\ldots,\delta a_{K}$ are the coefficients of the vector $\delta u$ in the basis of the $\Phi_{k}$, and note that 
\begin{equation*}
\begin{aligned}
\frac{\delta I}{\delta u} & = \left.\frac{d}{d\eps}\right|_{\eps=0} I(u+\eps \delta u)\\
& = \left(\lim_{\eps\to 0}\frac{I(u+\eps\delta u) - I(u)}{\eps}\right)\cdot \delta u\\
& = \left(\lim_{\eps\to 0}\frac{\bE_{P^{\eps}}[\psi(u+\eps \delta u)] - \bE_{P^{0}}[\psi(u)]}{\eps}\right)\cdot\delta u\\
\end{aligned}
\end{equation*} 
provided that the limit exists. The above iteration therefore is equivalent to
\begin{equation}\label{gd}
a_{k}^{(n+1)} = a_{k}^{(n)} + \tau_{n}\left.\frac{\partial I}{\partial a_{k}}\right|_{a_{k}=a_{k}^{(n)}}\,,
\end{equation}
with
\begin{equation}\label{dida}
\frac{\partial I}{\partial a_{k}} = \bE_{P^0}\left[\psi\int_{0}^{T}(\sigma^{-1}D(q_{s})\Phi_{k}(s))\cdot dB_{s} + \int_{0}^{T}u(s)\cdot\Phi_{k}(s)\,ds \right] \,,
\end{equation}
and $\bE_{P^0}[\cdot]$ being the expectation over all realizations of (\ref{eqn:langevin-1d}) with $\eps=0$ and $u=u^{(n)}$ being the current iterate and arbitrary initial conditions $x=(q,p)$, e.g., 
\[
\bE_{P^{0}}[\psi] = \int_{\R^{2n}}\bE_{P^{0}_{x}}[\psi]\,d\mu(x) \,,
\]
with $P_{x}^{0}$ denoting the corresponding path measure with deterministic initial condition, and $\mu$ is the prescribed initial distribution.
{A related result on computing the parameter sensitivity can be found
  in Ref.~\cite{warren2012malliavin}.}
Note that we can compute all partial derivatives $\partial I/\partial a_k$, $k=1,\ldots,K$ from just one ensemble of nonequilibrium paths of $(q_{t},p_{t})$. Further note that, in general, the updated coefficient $a^{(n+1)}$ and hence the updated control $u^{(n+1)}$ will depend on the distribution of the initial conditions at the $n$-th iteration stage; in particular, if the controls are computed on-the-fly, the controls will actually be feedback controls depending on the current state and the time elapsed since $t=0$.

@@

{
  We derived the first and second order
  response formulas for nonequilibrium molecular
dynamics (driven Langevin dynamics), which is a generalization of the
standard finite-time equilibrium linear response theory.
The novel nonequilibrium response theory
does not rely on any reversibility or stationarity assumption on the dynamics.
We validated the formula in numerical
experiments in comparison to brute-force nonequilibrium
simulations. There, we demonstrated that the nonequilibrium linear
response formula allows to extend the algorithmic use of linear
response theory to significantly stronger perturbations of the system
since it permits intermediate drivings, with which the effective perturbations
still lie in the linear range.
}
